\sys: Light-touch Data-Parallel Shell Processing

### 2020-10-06 paper/draft.tex
Pass over 4

\heading{Dataflow Regions}
The DFG model and associated transformations are applied only to program sub-ex\-pres\-sions that \sys deems would be safe to parallelizable---\ie ones that would not break sequential semantics, as long as the commands involved in these regions fall under the right classes.
%% respect to their sequential execution.
The search for these regions is guided by the shell language and the structure of a particular program.
These contain information about (i) fragments that can be executed independently, and (ii) barriers that are natural synchronization points. % for enforcing synchronization.
Consider this fragment:

@@

\noindent
The \ttt{cat} and \ttt{grep} commands execute independently (and concurrently) in the standard shell, but \ttt{sort} waits for their completion prior to start.
Both \ttt{cat f1 f2 | grep "foo" > f3} and \ttt{sort f3} are thus dataflow regions, each not extending beyond \ttt{&&}.\footnote{
  Can we make the commands \ttt{cmd1}, \ttt{cmd2}, \etc to ensure the readers understand this is driven by structure and not commands? \kk{Commands do matter a bit though. If a command is sideeffectful we don't know how to connect it in the dataflow graph and thus \sys would not parallelize it (as is written later in the translation pass).}}
% ---more precisely, they are maximal.

@@

\heading{Translation Pass}
%
%% \kk{It might be a good idea to change this subsection based on the new
%%   as-late-as-possible architecture where everything is wrapped and
%%   translated at the final moment. Ideally, this would become the main
%%   point of this subsection, as the translation algorithm is kind of
%%   redundant when having the distributable regions too.}
%
\sys's front-end performs a depth-first search on the AST of
the given shell program.  During this pass, it extends the
dataflow regions bottom-up, translating their
independent components to DFG nodes until a barrier
construct is reached. All AST subtrees not translatable to DFGs are kept as they are. The output of the translation pass is the
original AST where dataflow regions have been replaced with
DFGs and calls to \sys's runtime primitives~\sx{optimizer}.

To identify opportunities for parallelization, the translation pass extracts each command's  parallelizability class together with its inputs and outputs.
To achieve this for each command, it searches all its available annotations~\sx{ext} and resorts to conservative defaults if none is found.
If the command is in \sta or \pur, the translation pass initiates a dataflow region that is propagated up the tree.

@@

The two main shell abstractions are
  (i) streams, \ie files or pipes, that contain data, and
  (ii) commands that communicate through these streams.
%% \nv{quick solution: change all files to streams and make the equivalence explicit here}
\sys's DFG model represents commands as nodes and streams as edges.

We first introduce the notation used in this section.
For a set $D$, we write $D\kstar$ to denote the set of all finite words over $D$.
For words $x, y \in D\kstar$, we write $x \cdot y$ or $xy$ to denote their concatenation.
We write $\eps$ for the empty word.
We say that $x$ is a \emph{prefix} of $y$, and we write $x \leq y$, if there is a word $z$ such that $y = xz$.
The $\leq$ order is reflexive, antisymmetric, and transitive (\ie it is a partial order), and is often called the \emph{prefix order}.

@@

\subsection{Backend: From DFGs to Shell Scripts}
\label{backend}

After translating the dataflow regions of the input script to DFGs and applying optimization passes~\sx{ir:transformations},
  \sys translates all DFGs back into a shell script.
Nodes of the graph are instantiated with the commands and flags they represent, and edges are instantiated as named pipes.
A prologue in the script creates the necessary intermediate pipes, and a \ttt{trap} call takes care of cleaning things up in cases where the script aborts.
The script makes use of \sys-provided primitives available in the system \ttt{PATH} and described next.

@@

\tr{If there is time I can work out a formal definition and a proof
  sketch why this transformation preserves the output of the dataflow
  graph.}

>>>>

\heading{Dataflow Regions}
In order to apply the graph transformations that expose data parallelism, \sys first has to identify program sub-expressions that can be safely transformed to a dataflow graph, \ie sub-expressions that do not impose any scheduling or synchronization (e.g. by using \ttt{;}) and take a set of files as inputs and produce a set of files as outputs.
% The DFG model and associated transformations are applied only to program sub-ex\-pres\-sions that \sys deems would be safe to parallelizable---\ie ones that would not break sequential semantics, as long as the commands involved in these regions fall under the right classes.
%% respect to their sequential execution.
The search for these regions is guided by the shell language and the structure of a particular program.
These contain information about (i) fragments that can be executed independently, and (ii) barriers that are natural synchronization points. % for enforcing synchronization.
Consider this fragment:

@@

\noindent
The \ttt{cat} and \ttt{grep} commands execute independently (and concurrently) in the standard shell, but \ttt{sort} waits for their completion prior to start.
Both \ttt{cat f1 f2 | grep "foo" > f3} and \ttt{sort f3} are thus dataflow regions, each not extending beyond \ttt{&&}.\footnote{
  Can we make the commands \ttt{cmd1}, \ttt{cmd2}, \etc to ensure the readers understand this is driven by structure and not commands? 
  % \kk{Commands do matter a bit though. If a command is sideeffectful we don't know how to connect it in the dataflow graph and thus \sys would not parallelize it (as is written later in the translation pass).} 
  \kk{As we discussed we can add a small figure that shows the script, the AST, and the DFG.}}
% ---more precisely, they are maximal.

@@

\heading{Translation Pass}
%
%% \kk{It might be a good idea to change this subsection based on the new
%%   as-late-as-possible architecture where everything is wrapped and
%%   translated at the final moment. Ideally, this would become the main
%%   point of this subsection, as the translation algorithm is kind of
%%   redundant when having the distributable regions too.}
%
\sys's front-end performs a depth-first search on the AST of
the given shell program.  During this pass, it extends the
dataflow regions bottom-up, translating their
independent components to DFG nodes until a barrier
construct is reached. All AST subtrees not translatable to DFGs are kept as they are. The output of the translation pass is the
original AST where dataflow regions have been replaced with
DFGs. 
% and calls to \sys's runtime primitives~\sx{optimizer}.
% \kk{eurosys: I removed this since it is slightly inaccurate and readers don't know what the primitives are at that point.}

To identify opportunities for parallelization, the translation pass extracts each command's  parallelizability class together with its inputs and outputs.
To achieve this for each command, it searches all its available annotations~\sx{ext} and resorts to conservative defaults if none is found.
If the command is in \sta, \pur, or \npu, the translation pass initiates a dataflow region that is propagated up the tree.

@@

The two main shell abstractions are
  (i) streams, \ie files or pipes, that contain data, and
  (ii) commands that communicate through these streams.
%% \nv{quick solution: change all files to streams and make the equivalence explicit here}
\sys's DFG model represents commands as nodes and streams as edges.
We now introduce the notation used in throughout this section.
For a set $D$, we write $D\kstar$ to denote the set of all finite words over $D$.
For words $x, y \in D\kstar$, we write $x \cdot y$ or $xy$ to denote their concatenation.
We write $\eps$ for the empty word.
We say that $x$ is a \emph{prefix} of $y$, and we write $x \leq y$, if there is a word $z$ such that $y = xz$.
The $\leq$ order is reflexive, antisymmetric, and transitive (\ie it is a partial order), and is often called the \emph{prefix order}.

@@

\subsection{Backend: From DFGs to a Parallel Shell Script}
\label{backend}

After translating the dataflow regions of the input script to DFGs and applying the transformations~\sx{ir:transformations},
  \sys translates all DFGs back into a shell script.
Nodes of the graph are instantiated with the commands and flags they represent, and edges are instantiated as named pipes.
A prologue in the script creates the necessary intermediate pipes, and a \ttt{trap} call takes care of cleaning things up in cases where the script aborts.
% The script makes use of \sys-provided primitives available in the system \ttt{PATH} and described next.
% \kk{I removed this, since we describe them just below.}

@@
